exploratory

### 2024-06-26 exploratory.Rmd
trstee HMM fit & joint state analytics

---
title: "HMM fitting to data of participants returns in the Exposure experiment"
author: "Ismail Guennouni"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, include =F)
```

```{r}
library(tidyverse)
library(afex)
library(PairedData)
library(multcompView)
library(lsmeans)
library(depmixS4)
library(flextable)
library(grid)
library(gridExtra)
library(forcats)
library(ggsignif)
library(magick)
library(doParallel)
library(factoextra)

@@

## Fitting HMM model to participants data

Here we fit an HMM model to data generated by participants in both RTGs (pre and post manipulation) for both conditions. We specify that the transition between latent states depends on the investment received. In order to distinguish transitons by condition and Phase, we add a contrast variabel that has three levels (Pre-manipulation, post-control and post exposure). We compare models with different number of states and choose the one with the lowest BIC.

@@

# select columns of interest and also solve issue with the participant with two RS_scores by averaging them. 
data_HMM <- d_finished %>%
  dplyr::select(playerId, roundType, investment, returns, roundNum, gameNum.f, condition.f, RS_score, LPFS_score, investorState.f) %>% 
  filter(roundType == "trust", !is.na(roundNum), gameNum.f %in% c("pre", "post")) %>% 
  dplyr::select(-roundType) %>%
  mutate(roundNum = as.numeric(as.character(roundNum))) %>%
  mutate(inv_pct = investment/20, ret_pct_0 = ifelse(investment == 0, 0, returns/(3*investment))) %>%
  group_by(playerId, investment, returns, roundNum, gameNum.f, condition.f, LPFS_score, investorState.f, inv_pct, ret_pct_0) %>%
  summarize(RS_score = mean(RS_score, na.rm = TRUE)) %>%
  ungroup()

# Data for both pre and post games
data_HMM_all <- data_HMM %>% 
  arrange(playerId, gameNum.f, roundNum) %>%
  group_by(playerId,gameNum.f) %>% 
  dplyr::mutate(next_investment = lead(investment, default=0),
                full_ctrst = factor(ifelse(gameNum.f=="pre", 0,ifelse(condition.f=="Control",1,2)))) %>%
  ungroup() %>% 
  unique()

# FOCUS ON POST GAMES ONLY 
data_HMM_post <- data_HMM %>% 
  filter(gameNum.f == "post") %>% 
  group_by(playerId) %>% 
  dplyr::mutate(next_investment = lead(investment, default=0)) %>%
  ungroup() %>%
  unique()

<!-- Here we extend the depmixS4 package functionality for analyzing data using hidden Markov models (HMMs). It involves defining new response classes and their associated methods. -->
```{r}

@@

# Fitting HMM to ALL Participants data (Pre and Post)

```{r}
# 
# #| label: setup-parallel-3
# 
# # Set up the parallel backend
# num_cores <- 8
# #create the cluster
# my.cluster <- parallel::makeCluster(
#   num_cores,
#   type = "FORK"
#   )
# 
# #register it to be used by %dopar%
# doParallel::registerDoParallel(cl = my.cluster)
```

@@

# # Depmix simple Models
# priordat_all <- data_HMM_all |>
#   filter(roundNum==1)
# 
# tr_fdmod_all <- list()
# 
# max_ns <- 8
# 
# 
# set.seed(20240513)
# my_seeds <- sample(1000:50000, max_ns)
# 
# 
# # Fitting HMM models to data
# tr_fdmod_all <- foreach(state = 2:max_ns, .packages = c("depmixS4"), .combine = "c") %dopar% {
#   # Define depmix model with simple transition only depending on next_inv
#   print(state)
#   tr_simple <- depmix(returns ~ 1, data=data_HMM_all, nstates=state, transition=~ next_investment*full_ctrst, prior=~ investment, initdata=priordat_all, family=multinomial("mlogit"), ntimes=rep(15,206*2))
# 
#   mus <- (1:state)*(1/(state+1))
#   sigmas <- rep(1/(state+1), state)
#   if(state > 8) sigmas <- sigmas + .1
#   rModels_tr <- list()
#   for(s in 1:state) {
#     rModels_tr[[s]] <- list(vtdgaus(y=data_HMM_all$ret_pct_0, yield = 3*data_HMM_all$investment, pstart=c(mus[s],sigmas[s])))
#   }
# 
#   tr_dmod <- makeDepmix(response=rModels_tr, transition=tr_simple@transition, prior=tr_simple@prior, ntimes = rep(15,206*2), homogeneous=FALSE)
# 
#   tmp1 <- fit(tr_dmod, emcontrol=em.control(random.start = FALSE))
#   set.seed(my_seeds[state])
#   tmp2 <- multistart(tr_dmod, nstart=20)
#   if(!is.na(logLik(tmp2)) && logLik(tmp2) > logLik(tmp1)) {tmp1 <- tmp2}
#   tmp1
# }
# 
# # Stop the parallel backend
# stopImplicitCluster()
# 
# 
# save(tr_fdmod_all, file="data/tr_fdmod_all.RData", compression_level = 9)

@@

load("data/tr_fdmod_all.RData")

df_bics_all <- data.frame(nstate = 2:max_ns,
                          logLik = unlist(lapply(tr_fdmod_all, logLik)),
                          num_params= unlist(lapply(tr_fdmod_all, function(x) attr(logLik(x), which="df"))),
                          AIC = unlist(lapply(tr_fdmod_all, AIC)),
                          BIC = unlist(lapply(tr_fdmod_all, BIC)))
df_bics_all

# Here we choose BIC as best m
nstate_tr <- which.min(df_bics_all[,5]) + 1
nstate_tr

best_mod_all <- order_mod_vtdgaus(tr_fdmod_all[[nstate_tr-1]])
summary(best_mod_all)
```

@@

df_bics_all %>%
  mutate_if(is.numeric, format, digits=4, nsmall = 0, big.mark = ",") %>%
knitr::kable(booktabs = TRUE,
  col.names = c("#states", "log(lik)", "# pars", "AIC", "BIC"))
```

<!-- Function to get transiton probabilities from fitted HMM models -->
```{r}
get_trans_probs <- function(fmod, numCovIn, numCovTr) {
  # --- Parameters & Setup ---
  ns <- nstates(fmod) # Number of states in the HMM
  investment_values <- seq(0, 20, length = 100)
  # Create a data frame to store the transition probabilities.
  # Each row represents a combination of "from" state, "to" state, and "investment" value.
  trans_prob_all <- expand.grid(
    from = paste0("from State ", 1:ns),
    to = paste0("State ", 1:ns),
    investment = investment_values
  )
  trans_prob_all$prob_pre <- 0
  trans_prob_all$prob_ctrl <- 0
  trans_prob_all$prob_expo <- 0

  # --- Calculate Transition Probabilities ---
  for (from in 1:ns) { 
    start_idx <- ns * numCovIn + 1 + (from - 1) * numCovTr * ns
    end_idx <- start_idx + numCovTr * ns - 1
    pars <- matrix(getpars(fmod)[start_idx:end_idx], ncol = numCovTr)
    print(pars)

    # Initialize matrices to store transition probabilities before normalization
    y0 <- matrix(0, nrow = length(investment_values), ncol = ns) # Pre-intervention group
    y1 <- matrix(0, nrow = length(investment_values), ncol = ns) # Control group
    y2 <- matrix(0, nrow = length(investment_values), ncol = ns) # Exposed group

    for (to in 1:ns) {
      y0[, to] <- exp(pars[to, 1] + pars[to, 2] * investment_values)
      y2[, to] <- exp(pars[to, 1] + pars[to, 2] * investment_values + pars[to, 3] + pars[to, 5] * investment_values) # Exposed: Intercept + investment + group effect + interaction
      y1[, to] <- exp(pars[to, 1] + pars[to, 2] * investment_values + pars[to, 4] + pars[to, 6] * investment_values) # Control: Intercept + investment + group effect + interaction
    }

    # Normalize transition probabilities for each group to ensure they sum to 1 across all "to" states
    y0 <- y0 / rowSums(y0)
    y1 <- y1 / rowSums(y1)
    y2 <- y2 / rowSums(y2)

    for (to in 1:ns) {
      idx <- trans_prob_all$from == paste0("from State ", from) & 
             trans_prob_all$to == paste0("State ", to)
      trans_prob_all$prob_pre[idx] <- y0[, to]
      trans_prob_all$prob_ctrl[idx] <- y1[, to]
      trans_prob_all$prob_expo[idx] <- y2[, to]
    }
  }

  saveRDS(trans_prob_all, "data/trans_prob_all.RDS")
  return(trans_prob_all)
}
```

# Plotting the trustee HMM policy by state 
tr_policy <- list()
for(i in 1:nstates(best_mod_all)) {
  tr_policy[[i]] <- dens(vtdgaus(seq(0,1, length=61), yield=3*20, pstart=c(unlist(best_mod_all@response[[i]][[1]]@parameters))))  
}

@@

respplot <- data.frame(state = rep(1:nstates(best_mod_all), each=length(perc_return)),
                            perc_return = perc_return,
                            probability = unlist(tr_policy)) |>
  mutate(state = factor(paste0("s",state))) |> 
  ggplot(                 
       aes(x = perc_return,
           y = probability,
           fill = state)) +
  geom_bar(stat = "identity",
           position = "dodge") + 
  # labs(fill='Latent trustee state') +
  xlab("Percentage return") +
  theme_minimal() + 
  theme(legend.position = "right") + ggtitle("State conditional strategies")

##################################################################################
# Plotting Prior (initial) state probabilities (new)

@@

# Get parameters from model
params <- getpars(getmodel(best_mod_all, "prior"))

# Extract intercepts and investment coefficients, and set initial investment value
intercepts <- params[1:nstate_tr]
investments <- params[(nstate_tr + 1):(2 * nstate_tr)]
# set average investment of trustee in neutral state as value to calculate initial state probabilities in linear model
investment_value <- 9

@@

```{r,include=F}

trans_prob_all <- get_trans_probs(best_mod_all, numCovIn=2, numCovTr=6)

ctrl_prePost_plot <- ggplot(trans_prob_all,aes(x=investment)) + 
  geom_line(aes(y=prob_ctrl, colour = as.factor(to), linetype="Post-control")) + 
  geom_line(aes(y=prob_pre, colour = as.factor(to), linetype="Pre-control")) +
  facet_wrap(~from) + 
  ylim(c(0,1)) + 
  ggtitle("Transition function pre- and post-control group") + 
  labs(x = "Investment", y = "Transition probability", color='Transition to:') + 
  scale_linetype_manual(values = c("Post-control" = "solid", "Pre-control" = "dotted")) +
  theme_bw()

expo_prePost_plot  <- ggplot(trans_prob_all,aes(x=investment)) + 
  geom_line(aes(y=prob_expo, colour = as.factor(to), linetype="Post-exposure")) + 
  geom_line(aes(y=prob_pre, colour = as.factor(to), linetype="Pre-exposure")) +
  facet_wrap(~from) + 
  ylim(c(0,1)) + 
  ggtitle("Transition function pre- and post-exposure manipulation") + 
  labs(x = "Investment", y = "Transition probability", color='Transition to:', linetype = "Line Type") + 
  scale_linetype_manual(values = c("Post-exposure" = "solid", "Pre-exposure" = "dotted")) +
  theme_bw()

expo_prePost_plot 
ctrl_prePost_plot
```

```{r}
##############################################################################

#| label: fig-trustee-policy-plot
#| echo: false
#| fig.cap: HMM model of the trustees 
#| fig.align: center
#| fig.width: 12
#| fig.height: 6

# transplot <- plot_transitions(best_mod_all,numCovIn=2,numCovTr=2) + labs(x = "Investment", y = "Probability", color='Transition to') + theme_minimal() + theme(legend.position = "bottom")

gridExtra::grid.arrange(respplot, priorplot, expo_prePost_plot, ncol=3, layout_matrix=cbind(c(1,2),c(3,3),c(3,3)))
```

<!-- Get Trustee posterior states from model and add to data table -->

```{r, include = FALSE}
set.seed(20221010)
# mention we are using local decoding (refer to the book). 
predTrStatesAll <- posterior(best_mod_all, type="local")

#############
data_HMM_all$postTrState <-  factor(predTrStatesAll, levels= c(1,2,3,4,5),labels=c("State 1","State 2","State 3","State 4","State 5"))
############
```

#| label: fig-trustee-posterior-plot
#| echo: false
#| fig.cap: POsterior state of the trustee in eahc round by Condition and Phase
#| fig.align: center
#| fig.width: 8
#| fig.height: 12
postHMM <- ggplot(data_HMM_all) +
  geom_bar(aes(x = roundNum, group = postTrState, fill = postTrState),  position = position_stack(reverse = TRUE)) + 
  # facet_wrap(~factor(condition.f, levels = rev(levels(as.factor(condition.f))))) +  
  facet_wrap(~condition.f*gameNum.f)+
  labs(x = "Round", fill='Posterior trustee state') +
  theme_bw() +
  theme(legend.position = "bottom")

postHMM
```

<!-- Cluster joint latent states: -->

```{r}
# Load necessary libraries
library(factoextra)
library(hexbin)

# Assuming data_HMM_post is your dataframe

# Rename the levels of investorState and postTrState
data_HMM_all <- data_HMM_all %>%
  mutate(investorState = recode(investorState.f, 
                                "unhappy" = "1", 
                                "neutral" = "2", 
                                "happy" = "3"),
         postTrState = gsub("State ", "", postTrState))

# Define the desired order of joint states
desired_order <- c(
  "1-1", "1-2", "1-3", "1-4", "1-5",
  "2-1", "2-2", "2-3", "2-4", "2-5",
  "3-1", "3-2", "3-3", "3-4", "3-5"
)

# Ensure Joint_State and Next_Joint_State are factors with the specified order
data_HMM_all <- data_HMM_all %>%
  mutate(Joint_State = factor(paste(investorState, postTrState, sep = "-"), 
                              levels = desired_order, ordered = TRUE))

# Plot the frequency of each Joint_State
# Define a color palette based on the desired order
color_palette <- colorRampPalette(c("red","orange", "green", "cyan", "blue"))(length(desired_order))
state_pair_to_color <- setNames(color_palette, desired_order)

# Plot the frequency of each Joint_State with the assigned colors
ggplot(data_HMM_all, aes(x = Joint_State)) +
  geom_bar(aes(fill = Joint_State)) +
  scale_fill_manual(values = state_pair_to_color) +
  labs(title = "Frequency of Joint States",
       x = "Joint State",
       y = "Frequency") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#Plot joint states distribution by phase and condition 
ggplot(data_HMM_all, aes(x = Joint_State, fill = gameNum.f)) +
  geom_bar(position = position_dodge(width = 0.9)) +
  scale_fill_manual(values = c("lightblue", "blue"), # Example colors for different conditions
                    name = "Condition") +
  labs(title = "Frequency of Joint States by Condition",
       x = "Joint State",
       y = "Frequency") +
  facet_wrap(~condition.f)+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

# Create next joint state for transition analysis
data_HMM_all <- data_HMM_all %>%
  arrange(playerId, gameNum.f, roundNum) %>%
  group_by(playerId, gameNum.f) %>%
  mutate(Next_Joint_State = lead(Joint_State, order_by = roundNum)) %>%
  ungroup() %>%
  mutate(Next_Joint_State = factor(Next_Joint_State, levels = desired_order, ordered = TRUE))

# Write to CSV if needed (for reference or debugging)
write.csv(data_HMM_all, "data/data_HMM_all.csv")

# Calculate transition counts
transition_counts <- data_HMM_all %>%
  filter(!is.na(Next_Joint_State)) %>%
  count(Joint_State, Next_Joint_State)

# Convert counts to probabilities and reshape to wide format
transition_matrix <- transition_counts %>%
  group_by(Joint_State) %>%
  mutate(transition_prob = n / sum(n)) %>%
  ungroup() %>%
  dplyr::select(Joint_State, Next_Joint_State, transition_prob) %>%
  pivot_wider(names_from = Next_Joint_State, values_from = transition_prob, values_fill = list(transition_prob = 0))

# Ensure the column order in transition_matrix
transition_matrix <- transition_matrix %>%
  dplyr::select(Joint_State, all_of(desired_order))

# Visualize the transition matrix
transition_matrix_long <- transition_matrix %>%
  pivot_longer(cols = -Joint_State, names_to = "Next_Joint_State", values_to = "probability")

ggplot(transition_matrix_long, aes(x = Next_Joint_State , y = Joint_State, fill = probability)) +
  geom_tile() +
  scale_fill_gradient2(low = "white", high = "blue", mid = "lightblue", midpoint = 0.5, limit = c(0, 1), space = "Lab", name="Transition Probability") +
  theme_minimal() +
  labs(title = "Transition Matrix of Joint States", x = "Current Joint State", y = "Next Joint State") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```
```{r}
frequency_table <- data_HMM_all %>%
  count(Joint_State, condition.f, gameNum.f) %>%
  spread(gameNum.f, n, fill = 0) %>% 
  mutate(Difference = `post` - `pre`)

# Plot the differences
ggplot(frequency_table, aes(x = Joint_State, y = Difference, fill = condition.f)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  scale_fill_manual(values = c("lightblue", "blue"), # Example colors for different conditions
                    name = "Condition") +
  labs(title = "Difference in Frequency of Joint States by Condition (post - pre)",
       x = "Joint State",
       y = "Difference in Frequency") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

<!-- Clustering of whole time series using K-means  -->
```{r}

# Reshape the data
df_wide <- data_HMM_all %>%
  dplyr::select(playerId, gameNum.f,investorState, postTrState) %>%
  mutate(across(c(investorState, postTrState), as.numeric)) %>%
  group_by(playerId, gameNum.f) %>%
  mutate(id = row_number()) %>%
  pivot_wider(names_from = id, values_from = c(investorState, postTrState)) %>%
  ungroup()

# Separate playerId and gameNum.f columns for later use
playerIds <- df_wide$playerId
gameNums <- df_wide$gameNum.f

# Convert the rest of the data frame to a matrix
data_matrix <- df_wide %>% dplyr::select(-playerId, -gameNum.f) %>% as.matrix()

# Determine the optimal number of clusters using the elbow method
wss <- vector("numeric", length = 10)

for (k in 1:10) {
  set.seed(123)  # For reproducibility
  kmeans_result <- kmeans(data_matrix, centers = k, nstart = 25)
  wss[k] <- kmeans_result$tot.withinss
}

# Plot the elbow plot
plot(1:10, wss, type = "b", pch = 19, frame = FALSE,
     xlab = "Number of Clusters K",
     ylab = "Total Within-Clusters Sum of Squares",
     main = "Elbow Plot for Determining Optimal K")

<!-- # Perform k-means clustering with the optimal number of clusters -->
```{r}

set.seed(123)
final_kmeans <- kmeans(data_matrix, centers = 3, nstart = 25)

# Add cluster assignments to the original data
df_wide$cluster <- factor(final_kmeans$cluster)

# Combine playerId and gameNum.f with cluster assignments
df_wide_with_id <- df_wide %>% 
  mutate(playerId = playerIds, gameNum.f = gameNums)

# Join the cluster assignments back to the original data
df_clustered <- data_HMM_all %>% 
  left_join(df_wide_with_id %>% dplyr::select(playerId, gameNum.f, cluster), 
            by = c("playerId", "gameNum.f"))

df_clustered %>%
  group_by(cluster, Joint_State) %>%
  summarise(count = n(), .groups = 'drop') %>%
  group_by(cluster) %>%
  mutate(proportion = count / sum(count)) %>%
  ggplot(aes(x = factor(cluster), y = proportion, fill = factor(Joint_State))) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = state_pair_to_color) +
  theme_minimal() +
  labs(title = "Proportion of Joint States by Cluster",
       x = "Cluster",
       y = "Proportion",
       fill = "Joint State") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

df_clustered <- df_clustered %>%
  mutate(cluster_mood = case_when(
    cluster == 2 ~ "Joint Unhappy",
    cluster == 1 ~ "Joint Happy",
    cluster == 3 ~ "Joint Neutral",
    TRUE ~ "Other"  # This catches any other values, if they exist
  ) %>% factor(levels = c("Joint Unhappy", "Joint Neutral", "Joint Happy")))

# Save the results to a CSV file (optional)
write.csv(df_clustered, "clustered_data.csv", row.names = FALSE)

```{r}
# Calculate the mean path of joint states for each cluster over time
df_mean_path <- df_clustered %>%
  group_by(cluster_mood, roundNum) %>%
  summarize(meanInvestorState = mean(as.numeric(investorState), na.rm = TRUE),
            meanPostTrState = mean(as.numeric(postTrState), na.rm = TRUE)) %>%
  ungroup()

# Combine both states in a single plot using facets
df_mean_path_long <- df_mean_path %>%
  pivot_longer(cols = c(meanInvestorState, meanPostTrState), 
               names_to = "StateType", 
               values_to = "MeanState")

 # Define custom colors for each cluster
cluster_colors <- c("Joint Unhappy" = "red", "Joint Neutral" = "darkgreen", "Joint Happy" = "blue")

ggplot(df_mean_path_long, aes(x = roundNum, y = MeanState, color = cluster_mood)) +
  geom_line(size = 1.5) +
  geom_point(size = 2) +
  scale_color_manual(values = cluster_colors) +
  facet_wrap(~ StateType, scales = "free_y") +
  labs(title = "Mean Path of States by Cluster Over Time",
       x = "Time",
       y = "Mean State",
       color = "Cluster") +
  theme_minimal()

```{r}
cluster_means <- df_clustered %>%
  group_by(cluster_mood, roundNum) %>%
  summarise(across(c(investment, ret_pct_0), mean))

ggplot(cluster_means, aes(x = roundNum, y = investment, color = factor(cluster_mood))) +
  geom_line() +
  scale_color_manual(values = cluster_colors) +
  labs(x = "Round Number", y = "Average investment by cluster", color = "Cluster")

ggplot(cluster_means, aes(x = roundNum, y = ret_pct_0, color = factor(cluster_mood))) +
  geom_line() +
  scale_color_manual(values = cluster_colors) +
  labs(x = "Round Number", y = "Average % return by cluster", color = "Cluster")

# Load the patchwork library
library(patchwork)

# Assuming your cluster_means dataframe and cluster_colors vector are already defined
# First plot for investment
p1 <- ggplot(cluster_means, aes(x = roundNum, y = investment, color = factor(cluster_mood))) +
  geom_line() +
  scale_color_manual(values = cluster_colors) +
  labs(x = "Round Number", y = "Average investment by cluster", color = "Cluster") +
  theme_minimal()

# Second plot for % return
p2 <- ggplot(cluster_means, aes(x = roundNum, y = ret_pct_0, color = factor(cluster_mood))) +
  geom_line() +
  scale_color_manual(values = cluster_colors) +
  labs(x = "Round Number", y = "Average % return by cluster", color = "Cluster") +
  theme_minimal()

# Combine the two plots
combined_plot <- p1 / p2

# Print the combined plot
print(combined_plot)

```
```{r}

# Calculate cumulative rewards and investments for each game
cumulative_summary <- df_clustered %>%
  group_by(playerId, cluster_mood, gameNum.f) %>%
  summarise(
    cum_investor_reward = sum(returns - investment, na.rm = TRUE),
    cum_trustee_reward = sum(3*investment - returns, na.rm = TRUE),
    .groups = 'drop'
  )  %>%
  group_by(cluster_mood) %>%
  summarise(
    avg_investor_cum = mean(cum_investor_reward),
    avg_trustee_cum = mean(cum_trustee_reward)
  ) %>%
  ungroup()

# Print summary
print(cumulative_summary)

<!-- Correlate the Cluster with initial state of the trustee:  -->

```{r}
library(dplyr)
library(nnet)  # for multinomial logistic regression
library(broom)  # for tidy output
library(ggplot2)

# First, let's filter for the initial round and select relevant variables
df_initial <- df_clustered %>%
  filter(roundNum == 1) %>%
  dplyr::select(playerId,gameNum.f, condition.f, cluster_mood, investment, postTrState)

# Create a dataframe with the pre and post cluster moods for each player and condition
pre_post_data <- df_initial %>%
  dplyr::select(playerId, gameNum.f, condition.f, cluster_mood) %>%
  spread(key = gameNum.f, value = cluster_mood) %>%
  filter(!is.na(pre) & !is.na(post))

# Calculate the distribution of people between clusters for pre and post
distribution_pre <- pre_post_data %>%
  group_by(condition.f, pre) %>%
  summarise(count = n()) %>%
  mutate(time = "pre") %>%
  rename(cluster_mood = pre)

distribution_post <- pre_post_data %>%
  group_by(condition.f, post) %>%
  summarise(count = n()) %>%
  mutate(time = "post") %>%
  rename(cluster_mood = post)

# Combine the pre and post data
combined_distribution <- bind_rows(distribution_pre, distribution_post)

# Ensure the time column is a factor with levels in the correct order
combined_distribution$time <- factor(combined_distribution$time, levels = c("pre", "post"))

# Plot the evolution between pre and post
ggplot(combined_distribution, aes(x = time, y = count, group = cluster_mood, color = cluster_mood)) +
  geom_line(aes(linetype = cluster_mood)) +
  geom_point() +
  facet_wrap(~ condition.f, scales = "fixed") +
  theme_minimal() +
  labs(title = "Evolution of Cluster Moods Pre to Post by Condition",
       x = "Time",
       y = "Count",
       color = "Cluster Mood",
       linetype = "Cluster Mood") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
<!-- #Testing whether differences between pre and post cluster frequencies/proportions are statistically significant -->
```{r}
# Create a contingency table
contingency_table <- combined_distribution %>%
  filter(time == "pre") %>%
  inner_join(combined_distribution %>% filter(time == "post"), by = c("condition.f", "cluster_mood")) %>%
  dplyr::select(condition.f, cluster_mood, count.x, count.y) %>%
  pivot_wider(names_from = condition.f, values_from = c(count.x, count.y))

# Rename columns for clarity
colnames(contingency_table) <- c("cluster_mood", "Control_Pre", "Manipulation_Pre", "Control_Post", "Manipulation_Post")

# Print contingency table
print(contingency_table)

# Function to perform proportion test for each class
perform_prop_test <- function(pre_counts, post_counts, total_pre, total_post) {
  counts <- c(pre_counts, post_counts)
  totals <- c(total_pre, total_post)
  test_result <- prop.test(counts, totals)
  return(test_result)
}

# Total counts for Control and Manipulation groups
total_control_pre <- sum(contingency_table$Control_Pre)
total_control_post <- sum(contingency_table$Control_Post)
total_manipulation_pre <- sum(contingency_table$Manipulation_Pre)
total_manipulation_post <- sum(contingency_table$Manipulation_Post)

# Perform tests for Control group
control_results <- lapply(1:nrow(contingency_table), function(i) {
  perform_prop_test(contingency_table$Control_Pre[i], contingency_table$Control_Post[i], total_control_pre, total_control_post)
})

# Perform tests for Manipulation group
manipulation_results <- lapply(1:nrow(contingency_table), function(i) {
  perform_prop_test(contingency_table$Manipulation_Pre[i], contingency_table$Manipulation_Post[i], total_manipulation_pre, total_manipulation_post)
})

# Print results for Control group
cat("Control Group Proportion Tests:\n")
for (i in 1:length(control_results)) {
  cat(paste("\nCluster Mood:", contingency_table$cluster_mood[i]))
  print(control_results[[i]])
}

# Print results for Manipulation group
cat("\n\nManipulation Group Proportion Tests:\n")
for (i in 1:length(manipulation_results)) {
  cat(paste("\nCluster Mood:", contingency_table$cluster_mood[i]))
  print(manipulation_results[[i]])
}

```{r}
# Ensure cluster_mood is a factor
df_initial$cluster_mood <- factor(df_initial$cluster_mood)

# Fit the multinomial logistic regression model
model <- multinom(cluster_mood ~ investment + postTrState, data = df_initial)

# Summary of the model
summary(model)

# Get tidy output of the model
tidy_model <- tidy(model)
print(tidy_model)

# Visualize the effects
ggplot(df_initial, aes(x = investment, fill = cluster_mood)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~postTrState) +
  labs(title = "Distribution of Initial Investment by Cluster Mood and Trustee State",
       x = "Initial Investment Percentage",
       y = "Density") +
  theme_minimal()
```

### Key takeaways:

Higher initial investments slightly decrease the likelihood of being in Joint Neutral or Joint Happy clusters compared to Joint Unhappy.
Higher trustee states, particularly 3 and above, strongly increase the odds of being in Joint Neutral or Joint Happy clusters compared to Joint Unhappy.
The effect of trustee state is more pronounced for the Joint Happy cluster than for Joint Neutral.

These results suggest that the trustee's initial state is a stronger predictor of cluster membership than the initial investment amount, with higher trustee states associated with more positive joint outcomes.

>>>>

---
title: "exploratory"
author: "Ismail Guennouni"
date: "`r Sys.Date()`"
output: pdf_document
---

Here we fit an HMM to all trustee data in the exposure experiment, both pre/post and control/manipulation.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(afex)
library(PairedData)
library(multcompView)
library(lsmeans)
library(depmixS4)
library(flextable)
library(grid)
library(gridExtra)
library(forcats)
library(ggsignif)
library(magick)
library(doParallel)

@@

## Exploratory analyses to model individual games

@@

data_HMM <- d_finished %>%
  dplyr::select(playerId,roundType,investment,returns,roundNum,gameNum.f,condition.f, phase.f, high_RS, LPFS_score, RS_score) %>% 
  filter(roundType=="trust",!is.na(roundNum),gameNum.f %in% c("pre","post")) %>% 
  mutate(roundNum = as.numeric(as.character(roundNum))) %>%
  mutate(inv_scaled = as.vector(scale(investment))) %>% 
  mutate(next_investment = lead(investment, default=0)) %>%
  mutate(inv_pct = investment/20, ret_pct_0 = ifelse(investment==0,0,returns/(3*investment)),ret_pct_na = ifelse(investment==0,NA,returns/(3*investment)))
```

@@

#| label: setup-parallel-3

# Set up the parallel backend
num_cores <- 8
#create the cluster
my.cluster <- parallel::makeCluster(
  num_cores,
  type = "FORK"
  )

#register it to be used by %dopar%
doParallel::registerDoParallel(cl = my.cluster)
```

@@

# Depmix simple Models 
priordat <- data_HMM |>
  filter(roundNum==1)

tr_fdmod <- list()

max_ns <- 10

set.seed(20240513)
my_seeds <- sample(1000:50000, max_ns)

# Fitting HMM models to data 
tr_fdmod <- foreach(state = 2:max_ns, .packages = c("depmixS4"), .combine = "c") %dopar% {
  # Define depmix model with simple transition only depending on next_inv
  tr_simple <- depmix(returns ~ 1, data=data_HMM, nstates=state, transition=~ next_investment, prior=~ investment, initdata=priordat, family=multinomial("mlogit"), ntimes=rep(15,414))

  mus <- (1:state)*(1/(state+1))
  sigmas <- rep(1/(state+1), state)
  if(state > 8) sigmas <- sigmas + .1
  rModels_tr <- list()
  for(s in 1:state) {
    rModels_tr[[s]] <- list(vtdgaus(y=data_HMM$ret_pct_0, yield = 3*data_HMM$investment, pstart=c(mus[s],sigmas[s])))
  }

  tr_dmod <- makeDepmix(response=rModels_tr, transition=tr_simple@transition, prior=tr_simple@prior, ntimes = rep(15,414), homogeneous=FALSE)

  tmp1 <- fit(tr_dmod, emcontrol=em.control(random.start = FALSE))
  set.seed(my_seeds[state])
  tmp2 <- multistart(tr_dmod, nstart=20)
  if(!is.na(logLik(tmp2)) && logLik(tmp2) > logLik(tmp1)) {tmp1 <- tmp2}
  tmp1
}

# Stop the parallel backend
stopImplicitCluster()

save(tr_fdmod, file="data/tr_fdmod.RData", compression_level = 9)

@@

df_bics_tr <- data.frame(nstate = 2:max_ns,
                          logLik = unlist(lapply(tr_fdmod, logLik)),
                          unlist(lapply(tr_fdmod, function(x) attr(logLik(x), which="df"))),
                          AIC = unlist(lapply(tr_fdmod, AIC)),
                          BIC = unlist(lapply(tr_fdmod, BIC)))
df_bics_tr

# Here we choose BIC as best m
nstate_tr <- which.min(df_bics_tr[,5]) + 1

best_mod_tr_all <- order_mod_vtdgaus(tr_fdmod[[nstate_tr-1]])
summary(best_mod_tr_all)
```

@@

df_bics_tr %>%
  mutate_if(is.numeric, format, digits=4, nsmall = 0, big.mark = ",") %>%
knitr::kable(booktabs = TRUE,
  col.names = c("#states", "log(lik)", "# par", "AIC", "BIC"))
```

# Plotting the trustee HMM policy by state 
tr_policy <- list()
for(i in 1:nstates(best_mod_tr_all)) {
  tr_policy[[i]] <- dens(vtdgaus(seq(0,1, length=61), yield=3*20, pstart=c(unlist(best_mod_tr_all@response[[i]][[1]]@parameters))))  
}

@@

respplot <- data.frame(state = rep(1:nstates(best_mod_tr_all), each=length(perc_return)),
                            perc_return = perc_return,
                            probability = unlist(tr_policy)) |>
  mutate(state = factor(paste0("s",state))) |> 
  ggplot(                 
       aes(x = perc_return,
           y = probability,
           fill = state)) +
  geom_bar(stat = "identity",
           position = "dodge") + 
  # labs(fill='Latent trustee state') +
  xlab("Percentage return") +
  theme_minimal() + 
  theme(legend.position = "right") + ggtitle("State conditional strategies")

# Plotting Prior (initial) state probabilities
# priorplot <- data.frame(state=factor(paste0("s",1:nstate_tr)), prob=getpars(getmodel(best_mod_tr_all, "prior"))) |>
#   ggplot(aes(x = state, y = prob, fill=state)) +
#   geom_bar(stat = "identity", width = 0.5) +
#   xlab("state") +
#   ylab("prior probability") +
#   ggtitle("Initial state probability") +
#   theme_minimal()

@@

# Assuming best_mod_tr_all is your fitted model
params <- getpars(getmodel(best_mod_tr_all, "prior"))

@@

# Plotting Transition function between states
plot_transitions <- function(fmod,numCovIn,numCovTr) {

  ns <- nstates(fmod)

  trans_prob <- data.frame(
    from = rep(1:ns, each=80*ns),
    to = rep(1:ns, each=80),
    inv = seq(0,20,length=80),
    prob = 0
  )

  y <- matrix(0.0,ncol=ns, nrow=80)

  for(from in 1:ns) {
    #pars <- matrix(getpars(fmod)[seq(ns + 1 + (from - 1)*2*ns, ns + 1 + (from - 1)*2*ns + 2*ns - 1)],ncol=2)
    pars <- matrix(getpars(fmod)[seq(ns*numCovIn + 1 + (from - 1)*numCovTr*ns, ns*numCovIn + 1 + (from - 1)*numCovTr*ns + numCovTr*ns - 1)], ncol=numCovTr)
    paste0("from state: ", from)
    print(pars)
    for(to in 1:ns) {
      x <- trans_prob[trans_prob$from == from & trans_prob$to == to, "inv"]
      y[,to] <- exp(pars[to,1] + pars[to,2]*x)
    }
    y <- y/rowSums(y)
    for(to in 1:ns) {
      trans_prob$prob[trans_prob$from == from & trans_prob$to == to] <- y[,to]
    }
  }
